Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-06-15 analysis/paper/paper.Rmd
Final edits before sending the text off to read-through

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/jas_reports.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as raw-material variability and excavation and classification practices, the sites chosen for analysis have all been excavated relatively recently, and have a constrained geographical distribution. Reaching the conclusion that these measures are applicable in a Norwegian Mesolithic setting. While this certainly holds potential for larger scale comparative studies, it also sets the stage for a more targeted and rigorous model evaluation outside an exploratory setting.

keywords: |
  Mesolithic Norway; Lithic assemblages; Multivariate statistics;
---

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [e.g. @clark2017], which has been employed in other contexts to align properties of the assemblages with land-use and mobility patterns.

@@

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @glørstad2011; @eigeland2015, 370]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern searching approach does constitute a limited inferential framework [@clark2009], as it involves a post hoc accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses. This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].

@@

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2013; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181]. In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following [@Bicho2020] the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative as compared to the CA [cf. @baxter1994, 100].

A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g. @breivik2016; @Bicho2020]. For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the previously mentioned Frison effect. Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017]. This pertains for example to the impact of bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014[124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. The use of metarhyolite for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b]. In sum, the findings for the earliest sites are in large part in line with previous research [e.g. @jaksland2014b; @breivik2018; @damlien2018b].

@@

Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed. These are displayed in the correlation matrix in in Figure \@ref(fig:cor-plot). Figure \@ref(fig:pca-analysis) displays the resulting PCA. There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot. The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure \@ref(fig:var-contrib)). Almost orthogonal to this is the strong negative correlation between relative frequency of chips and blanks. While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time. However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation. Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot)), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification. This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material. While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern. Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well. That is not to say that cores may not be indicative or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].

@@

```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA. The mean age of the sites has also been included to visualise overall temporal trends. Cells below the diagonal display the bivariate distributions with a fitted OLS-regression. The cells above the diagonal display and are coloured by the corresponding Pearsons correlation coefficient.", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate(
         "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3), .keep = "unused") %>% 
  relocate("mean_date",
    "log10(lithic_dens)", "nonflint_freq^(1/3)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

@@

```{r simplot, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint. The temporal uncertainty is handled by means of a simulation approach where the site ages are drawn from their repspective age determination probability density functions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation run is plotted with some transparency. Sites mentioned in the text are given colour.", fig.height = 14, fig.width = 12}

@@

The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component. One possible implication of this could be that the aggregation of artefact categories might have been overly conservative. However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'. The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern. Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear. The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.

@@

The results of the CA align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories. These tendencies are already well-established when it comes to the formal tool types, but have been given less focus in light of entire assemblages. Precisely what behavioural implication the development in the occurrences of the tool and debitage categories have are less clear, but appears to follow a different and more complex development over time than that of curation, as operationalised here.

The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns is the main driver behind this empirical pattern --- in line with the framework of @barton2011. Another perspective would be that this is not surprising at all [cf. @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different. If its applicability and relevance for this setting is accepted, the relationship does hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond. Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with Mesolithic south-eastern Norway. The explicit quantification does, however, offer the possibility of more formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.

Having reached the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting can only constitute a first analytical step. As @elster2015[12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached. This, however, can only constitute the absolute minimum of what is required of a proposed explanation. Subsequent steps should be to probe and challenge this explanatory framework.

>>>>

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Department of Archaeology, History and Conservation
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/jas_reports.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages. The findings pertaining to the artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research. Furthermore, the chronological trends associated with variables taken from the so-called whole assemblage behavioural index, originally developed for characterising Paleolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside an exploratory setting, which might ultimately yield a powerful comparative tool for the analysis of Mesolithic assemblages.

keywords: |
  Mesolithic Norway; Lithic assemblages; Mobility; Multivariate statistics
---

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called whole assemblage behavioural index [e.g. @clark2017], which has been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.

@@

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades (width > 8mm and < 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @eigeland2015, 370; @glørstad2011]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium, following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data --- data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses. This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].

@@

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181]. In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [cf. @baxter1994, 71--77, 103].

A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g. @Bicho2020; @breivik2016a]. For the assemblage data itself this especially pertains to diversity in tool-types [see also @canessa2021], which has been omitted in light of the previously mentioned Frison effect. Some site specific aspects such as number of features has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017]. This pertains for example to the impact of bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have been relatively frequent on several of the sites treated here [@jørgensen2017; @darmark2018e].

@@

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that its assemblage is dominated by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014[124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. The use of metarhyolite for the production of axes is present at other contemporary sites as well, but is evidently not as prominent a part of these assemblages [@jaksland2014b]. In sum, the findings for the earliest sites are in large part in line with previous research [e.g. @breivik2018; @jaksland2014b; @damlien2018b].

@@

Moving on to the PCA of measures that have been linked to mobility, some of the variables with severely skewed distributions were initially transformed. These are displayed in the correlation matrix in Figure \@ref(fig:cor-plot). Figure \@ref(fig:pca-analysis) displays the resulting PCA. There is a general temporal transition from older to younger sites from the upper left to the bottom right of the plot. The second dimension is mainly defined by a negative correlation between the VDL and RFSL (Figure \@ref(fig:var-contrib)). Almost orthogonal to this is the strong negative correlation between relative frequency of chips and blanks. While there is a slight tendency for blanks to be more associated with younger sites, frequency of chips appears to be largely independent of time. However, the almost suspiciously strong negative correlation between chips and blanks can perhaps have a practical explanation. Seeing as the frequency of non-flint material is positively correlated with blanks and negatively correlated with chips (Figure \@ref(fig:cor-plot)), one explanation to this pattern could be that smaller non-flint pieces are simply more difficult to identify and separate from naturally fragmented stone during excavation and classification. This could conceivably have led to an over-representation of blanks as compared to chips in assemblages with a high proportion of non-flint material. While this is not necessarily the entire explanation behind the relationship, this does make it difficult to place much analytical weight on this pattern. Relative frequency of cores is not especially impactful in the PCA, and appears to be independent the temporal dimension as well. That is not to say that cores may not be indicative or related to mobility patterns, but to get at this may require further analysis beyond their simple classification as cores [@kitchel2021].

@@

```{r cor-plot, fig.cap = "Correlation matrix showing transformation of skewed variables for the PCA. The mean age of the sites has also been included to visualise overall temporal trends. Cells below the diagonal display the bivariate distributions with a fitted OLS-regression. The cells above the diagonal display and are coloured by the corresponding Pearson's correlation coefficient.", fig.height = 9.5, fig.width = 10}
# Transform the skewed variables - sqrt() is used where there are zero values,
# and seems to provide an adequate result.
lithicstr <- lithics2 %>%
  # Transform skewed variables and reorder
  mutate(
         "log10(lithic_dens)" = log10(lithic_dens),
         "log10(secondary_freq)" = log10(secondary_freq),
         "core_freq^(1/3)" = core_freq^(1/3),
         "nonflint_freq^(1/3)" = nonflint_freq^(1/3), .keep = "unused") %>% 
  relocate("mean_date",
    "log10(lithic_dens)", "nonflint_freq^(1/3)", "blanks_freq", "log10(secondary_freq)",
    "chip_freq", "core_freq^(1/3)" 
  )

@@

```{r simplot, fig.cap = "Temporal variation in the curation index for A) All lithics, and B) Flint. The temporal uncertainty is handled by means of a simulation approach where the site ages are drawn from their respective age determination probability density functions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation run is plotted with some transparency. Sites mentioned in the text are given colour.", fig.height = 14, fig.width = 12}

@@

The results of the CA does appear to align well with previous research, and the employed artefact categories are clearly capturing a temporal component. One possible implication of this close correspondence could be that the aggregation of artefact categories might have been overly conservative. However, it is also evidently clear, in the words of @kruskal1971 [22], that 'time is not the only dimension'. The results of the CA do most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, which is also made evident by some of significant deviances from the overall pattern. Unpicking and aligning these patterns with any specific behavioural and technological dimensions using the coarse CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural and social significance of the employed units of analysis unclear. The results of the CA can, however, be used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions to nuance or explain discrepancies in this data.

@@

The results of the CA align more or less with results of previous research in south-eastern Norway. This would indicate that in general, meaningful chronological patterning is associated with the employed artefact categories. These tendencies are already well-established when it comes to the formal tool types and some debitage categories, but have been given less focus in light of entire assemblages. Precisely what behavioural implication the development in the occurrences of the tool and debitage categories have are less clear, but appears to follow a different and more complex development over time than that of curation, as operationalised here.

The temporal trends associated with the curation index corresponds surprisingly well with trajectories of cultural development previously suggested in the literature, and does therefore, in my view, suggest that shifts in land-use and mobility patterns are the main drivers behind this empirical pattern --- in line with the framework of @barton2011. Another perspective would be that this is not surprising at all [cf. @kuhn2015, 14], and that the previously demonstrated relevance of these measures across a wide range of contexts points to their pervasive relevance for the organisation of lithic technology, and, therefore, that there should be little reason to think Mesolithic south-eastern Norway should be any different. However, the conclusion that these these measures apply to and appear to capture the dimensions of interest in a controlled empirical setting by means of an exploratory analysis can only constitute a first analytical step. As @elster2015[12] has pointed out, the human mind seems to have a propensity to settle for an explanation that *can* be true, as soon as this has been reached. This, however, can only constitute the absolute minimum of what is required of a proposed explanation. Subsequent steps should be to probe and challenge this explanatory framework, also in light of alternative hypotheses. The relationship does nonetheless hold great potential for large scale comparative studies in Mesolithic Scandinavia and beyond. Furthermore, the temporal trends associated with the curation index was here simply narratively associated with the most immediate chronological trends emphasised in the literature concerned with the Mesolithic of south-eastern Norway. The explicit quantification does, however, offer the possibility to conduct formal comparisons with a wide range of environmental, demographic and cultural dimensions across multiple scales of analysis.
